---
title: "Exercise 03: EDA-challenge"
author: "John Hinkle"
format: html
editor: visual
---

## Exercise 03

### Load the necessary packages

```{r}
library(tidyverse)
library(ggplot2)
```

### Load in the data file

```{r}
f <- "/Users/jeh5663/Documents/exercise-03/data-wrangling.csv"
d <- read_csv(f, col_names = TRUE)
```

### 1) Create the 'BSD' (body size dimorphism) variable

```{r}
bmmm <- d$Body_mass_male_mean # male body mass
bmfm <- d$Body_mass_female_mean # female body mass
bsd <- bmmm / bmfm # calculate body size dimorphism
d <- mutate(d, BSD = bsd) # add body size dimorphism to data set
```

### 2) Create the 'sex_ratio' variable

```{r}
females <- d$AdultFemale # quantity of females
males <- d$AdultMales # quantity of males
sr <- females / males # calculate sex ratio
d <- mutate(d, sex_ratio = sr) # add sex ratio to data set
```

### 3) Create 'DI' variable
DI or defensibility index is the ratio of day range length to the diameter of the home range

```{r}
day_range_length <- d$DayLength_km # day range length
home_range_area <- d$HomeRange_km2 # day range area
home_range_diamater <- (sqrt(home_range_area/pi))*2 # calculate day range diameter
di <- day_range_length / home_range_diamater # calculate defensibility index
d <- mutate(d, DI = di) # add defensibility index to data set
```

### 4) Plot of day range length vs time spent moving

Overall plot

```{r}
ggplot(d, aes(x=Move, y=DayLength_km)) + 
  geom_point(color="#BF5700") # plot points in burnt orange (for fun)

ggplot(d, aes(x=Move, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  scale_x_log10() +
  scale_y_log10()
```

Plots for each taxonomic Family

```{r}
ggplot(d, aes(x=Move, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  facet_wrap(~Family)
```

### 5) Plot of day range length vs group size

Overall

```{r}
ggplot(d, aes(x=MeanGroupSize, y=DayLength_km)) + 
  geom_point(color="#BF5700")
```

Plots for each taxonomic Family

```{r}
ggplot(d, aes(x=MeanGroupSize, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  facet_wrap(~Family)
```

### 6) Plot of canine size dimorphism vs body size dimorphism

Overall

```{r}
ggplot(d, aes(x=BSD, y=Canine_Dimorphism)) + 
  geom_point(color="#BF5700")
```

Plots for each taxonomic Family

```{r}
ggplot(d, aes(x=BSD, y=Canine_Dimorphism)) + 
  geom_point(color="#BF5700") +
  facet_wrap(~Family)
```

### 7a) Create diet_strategy variable
Frugivore, Folivore, or Omnivore

```{r}
d <- mutate(d, diet_strategy = ifelse(Fruit >= 50, "frugivore", ifelse(Leaves >= 50, "folivore",
    ifelse(Fruit < 50 & Leaves < 50, "omnivore", NA))))
```

### 7b) Create a boxplot of group size vs diet_strategy

```{r}
ggplot(drop_na(d, diet_strategy), aes(x= diet_strategy, y= MeanGroupSize)) +
  geom_boxplot(color="#BF5700")
```

### 8) {dplyr} and pipe exercise

```{r}
print (d <- mutate(d, Binomial = paste(Genus, Species, sep = " ")) |> # create 'Binomial' variable and add to df
  select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) |> # trim df
  group_by(Family) |> # Group by 'Family'
  summarize(n_cases = n(), avgBS = mean(Brain_Size_Species_Mean, na.rm = TRUE),
    avgMBM = mean(Body_mass_male_mean, na.rm = TRUE)) |> # calculate average brain size and average male body mass
  arrange(avgBS)) # sort by increasing brain size
```

