---
title: "Exercise 03: EDA-challenge"
author: "John Hinkle"
format: html
editor: visual
---

## Exercise 03

### Load the necessary packages

```{r}
library(tidyverse)
library(ggplot2)
```

### Load in the data file

```{r}
f <- "/Users/jeh5663/Documents/exercise-03/data-wrangling.csv"
d <- read_csv(f, col_names = TRUE)
```

### 1) Create the 'BSD' (body size dimorphism) variable

```{r}
bmmm <- d$Body_mass_male_mean # male body mass
bmfm <- d$Body_mass_female_mean # female body mass
bsd <- bmmm / bmfm # calculate body size dimorphism
d <- mutate(d, BSD = bsd) # add body size dimorphism to data set
```

### 2) Create the 'sex_ratio' variable

```{r}
females <- d$AdultFemale # quantity of females
males <- d$AdultMales # quantity of males
sr <- females / males # calculate sex ratio
d <- mutate(d, sex_ratio = sr) # add sex ratio to data set
```

### 3) Create 'DI' variable

DI or defensibility index is the ratio of day range length to the diameter of the home range

```{r}
day_range_length <- d$DayLength_km # day range length
home_range_area <- d$HomeRange_km2 # day range area
home_range_diamater <- (sqrt(home_range_area/pi))*2 # calculate day range diameter
di <- day_range_length / home_range_diamater # calculate defensibility index
d <- mutate(d, DI = di) # add defensibility index to data set
```

### 4) Plot of day range length vs time spent moving

Overall plot

```{r}
ggplot(d, aes(x=Move, y=DayLength_km)) + 
  geom_point(color="#BF5700") # plot points in burnt orange (for fun)
```

Let's try log transforming both variables and plotting them. This yields a more linear relationship. The relationship is positive, which suggests that primates that move farther tend to spend more time moving.

```{r}
ggplot(d, aes(x=Move, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  scale_x_log10() +
  scale_y_log10()
```

Plots for each taxonomic Family

```{r}
ggplot(d, aes(x=Move, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  facet_wrap(~Family)
```

Now, log transform each variable to see if any linear trends appear.

```{r}
ggplot(d, aes(x=Move, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~Family)
```

Based on both of these plots, it the linear trend breaks down when you plot by taxonomic Family. There is a semblance of a positive linear relationship for Family Atelidae. This family of New World monkeys includes spider and howler monkeys.

### 5) Plot of day range length vs group size

Overall. There is a semblance of a positive linear trend, but lets transform the variables and re-plot.

```{r}
ggplot(d, aes(x=MeanGroupSize, y=DayLength_km)) + 
  geom_point(color="#BF5700")
```

Log transformed. The positive linear relationship *appears* stronger. This supports the idea that species that travel further tend to live in larger groups. Perhaps those larger groups feel safer, allowing them to travel further.

```{r}
ggplot(d, aes(x=MeanGroupSize, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  scale_x_log10() +
  scale_y_log10()
```

Plots for each taxonomic Family, untransformed

```{r}
ggplot(d, aes(x=MeanGroupSize, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  facet_wrap(~Family)
```

After transforming both variables, it seems like the positive linear trend also shows up for several taxonomic families. This trend is observed for New World families (ex. Atelidae) and Old World (ex. Cercopithecidae)

```{r}
ggplot(d, aes(x=MeanGroupSize, y=DayLength_km)) + 
  geom_point(color="#BF5700") +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~Family)
```

### 6) Plot of canine size dimorphism vs body size dimorphism

Overall

```{r}
ggplot(d, aes(x=BSD, y=Canine_Dimorphism)) + 
  geom_point(color="#BF5700")
```

Plots for each taxonomic Family

```{r}
ggplot(d, aes(x=BSD, y=Canine_Dimorphism)) + 
  geom_point(color="#BF5700") +
  facet_wrap(~Family)
```

### 7a) Create 'diet_strategy' variable

Frugivore, Folivore, or Omnivore

```{r}
d <- mutate(d, diet_strategy = ifelse(Fruit >= 50, "frugivore", ifelse(Leaves >= 50, "folivore",
    ifelse(Fruit < 50 & Leaves < 50, "omnivore", NA))))
```

### 7b) Create a box plot of group size vs diet strategy

```{r}
ggplot(drop_na(d, diet_strategy), aes(x= diet_strategy, y= MeanGroupSize)) +
  geom_boxplot(color="#BF5700")
```

### 8) {dplyr} and pipe exercise

```{r}
print (d <- mutate(d, Binomial = paste(Genus, Species, sep = " ")) |> # create 'Binomial' variable and add to df
  select(Binomial, Family, Brain_Size_Species_Mean, Body_mass_male_mean) |> # trim df
  group_by(Family) |> # Group by 'Family'
  summarize(n_cases = n(), avgBS = mean(Brain_Size_Species_Mean, na.rm = TRUE),
    avgMBM = mean(Body_mass_male_mean, na.rm = TRUE)) |> # calculate average brain size and average male body mass
  arrange(avgBS)) # sort by increasing brain size
```
